### 2022-09-27 main/chapter_context.tex
some improvements to SM; some misc tweaks

\begin{singlespacing}
\chapter{Context: theory and experiment}
\label{chapter:context}
%
\begin{epigraphs}
\qitem{%
The idea that theorems follow from the postulates does not correspond to simple
observation.
If the Pythagorean theorem were found to not follow from the postulates, we
would again search for a way to alter the postulates until it was true.
Euclid's postulates came from the Pythagorean theorem, not the other way.}%
{Richard~Hamming,
\textit{The Unreasonable Effectiveness of Mathematics},
1980~\cite{hamming1980unreasonable}}
% \qitem{%
% One geometry can not be more true than another; it can only be more convenient.
% Geometry is not true, it is advantageous.}%
% {Robert~M.~Pirsig,
% \textit{Zen and the Art of Motorcycle Maintenance},
% 1974~\cite{pirsig1999zen}}
\end{epigraphs}
\end{singlespacing}

\noindent Theory is needed to understand experimental data,
and data are needed to improve theoretical models.
For the experimental work of this thesis to be better understood, this chapter
introduces its theoretical and experimental contexts;
Section~\ref{sec:context_sm_and_beyond} presents the current culmination of
theoretical particle physics in the Standard Model, its limitations, and its
possible future directions including supersymmetric ideas, and
Section~\ref{sec:context_atlas} then describes the \atlas\ detector, with whose
data we hope refine the Standard Model into its improved future forms.

\begin{singlespacing}
\section{The Standard Model and beyond}
\label{sec:context_sm_and_beyond}
%
\begin{epigraphs}
\qitem{%
In fact, it is a defensible conjecture that no
theory can be complete and inviolable that emerges from finitely equipped minds
and survives finitely scoped experiment, which are the twin rickety foundations
for all theories.%
}%
{James~D.~Wells,
\textit{Discovery Beyond the Standard Model of~Elementary~Particle~Physics},
2019~\cite{wells2020discovery}}
\end{epigraphs}
\end{singlespacing}

\noindent Our best current understanding of Nature's fundamental objects in
their observable behaviours as particles is encoded in the Standard Model,
which uses solid mathematical foundations to accurately describe the physics
of particle interactions across a broad domain.
But that domain is bounded, so the Standard Model is not an ultimate theory of
everything, and work continues to expand that boundary on both theoretical and
experimental fronts.
The current Standard Model is described in Section~\ref{sec:context_sm},
along with its limitations and possible future directions.

@@

\subsection{The Standard Model}
\label{sec:context_sm}
% \TODO{particle content figure/tables}
The current Standard Model has three generations of fermions
that interact through forces mediated by vector bosons
and the scalar Higgs boson.
The strong force is mediated by
Massless gluons $g$ mediate the strong force,
the massless photon $\gamma$ mediates the electromagnetic force,
and very massive $W^+$, $W^-$, and $Z$ bosons weadiate the weak force.
\TODO{}

@@

Mathematically, the Standard Model is described by a quantized gauge field
theory with the symmetry group
$SU\!(3)_C \times SU\!(2)_L \times U\!(1)_Y$,
meaning that its Lagrangian is invariant under local operations of group
elements on its fields in certain representations.

Gluons correspond to the eight generators of $SU\!(3)_C$, whose subscript $_C$
indicates that they operate on colour charge triplets.

Unified electroweak interactions come from $SU\!(2)_L$ and $U\!(1)_Y$, whose
four bosons $W^1$, $W^2$, $W^3$, and $B$ respectively would be massless
without the Higgs mechanism, which inserts four scalar fields and
`spontaneously breaks' the gauge symmetry with a potential that moves the
ground state of the model to a gauge-asymmetric configuration.
From this new ground state, linear combinations of the electroweak fields
acquire masses while ``eating up'' three of the four scalar fields to yield
the three weak bosons and the photon:
$(W^1, W^2)$ mix into $(W^+, W^-)$, and $(W^3, B)$ mix into $(Z, \gamma)$.
The fourth and final scalar field is the Higgs boson $h$~\cite{
rubakov1999classical,
cottingham2007greenwood
}.

\TODO{(Higgs) mechansim~\cite{
higgs1964broken,
englert1964broken
}}

@@

\begin{singlespacing}
\section{The \atlas\ Experiment}
\label{sec:context_atlas}
%
\begin{epigraphs}
\qitem{%
if observing outer space gives us a view of the past, observing inner space
would surely give us a glimpse into the future - would be interesting if NASA
made a telescope for that%
}%
{Ken~M,
\textit{Comment: Yahoo! News},
2012~\cite{kenm2012inner}}
\end{epigraphs}
\end{singlespacing}

\noindent Towards explaining the main result of this thesis,
a search for effects from supersymmetric models
in data from the \atlas\ detector
on the Large Hadron Collider (LHC)~\cite{
atlas2022searches,
atlas2008experiment,
lhc2008machine
},
we first cover some relevant facts about the design and motivations
behind \atlas.

>>>>

\begin{singlespacing}
\chapter{Context: theory and experiment}
\label{chapter:context}
%
\begin{epigraphs}
\qitem{%
The idea that theorems follow from the postulates does not correspond to simple
observation.
If the Pythagorean theorem were found to not follow from the postulates, we
would again search for a way to alter the postulates until it was true.
Euclid's postulates came from the Pythagorean theorem, not the other way.}%
{Richard~Hamming,
\textit{The Unreasonable Effectiveness of Mathematics},
1980~\cite{hamming1980unreasonable}}
\qitem{%
if observing outer space gives us a view of the past, observing inner space
would surely give us a glimpse into the future - would be interesting if NASA
made a telescope for that%
}%
{Ken~M,
\textit{Comment: Yahoo! News},
2012~\cite{kenm2012inner}}
\end{epigraphs}
\end{singlespacing}

\noindent Theory is needed to understand experimental data,
and data are needed to improve theoretical models.
To clarify the experimental work of this thesis, this chapter introduces its
theoretical and experimental contexts;
Section~\ref{sec:context_sm_and_beyond} presents the current culmination of
theoretical particle physics in the Standard Model, its limitations, and its
possible future directions including supersymmetric ideas, and
Section~\ref{sec:context_atlas} then describes the \atlas\ detector, with whose
data we hope refine the Standard Model into its improved future forms.

\section{The Standard Model and beyond}
\label{sec:context_sm_and_beyond}
\noindent Our best current understanding of Nature's fundamental objects in
their observable behaviours as particles is encoded in the Standard Model,
which uses solid mathematical foundations to accurately describe the physics
of particle interactions across a broad domain.
But that domain is bounded, so the Standard Model is not an ultimate theory of
everything, and work continues to expand that boundary on both theoretical and
experimental fronts.
The current Standard Model is described in Section~\ref{sec:context_sm},
along with its limitations and possible future directions.

@@

\subsection{The Standard Model}
\label{sec:context_sm}
\TODO{particle content figure/tables}

The current Standard Model features three generations of fermions,
which interact through forces mediated by vector bosons
and the scalar Higgs boson.
The strong force is mediated by
Massless gluons $g$ mediate the strong force,
the massless photon $\gamma$ mediates the electromagnetic force,
and very massive $W^+$, $W^-$, and $Z$ bosons weadiate the weak force.

Gauge field theory describes this standard model with the symmetry group
$SU\!(3)_C \times SU\!(2)_L \times U\!(1)_Y$,
meaning that its Langrangian is invariant under local operations of the group
on its fields in certain representations.
Gluons correspond to the eight generators of $SU\!(3)_C$, whose subscript $_C$
indicates that they operate on colour charge triplets.
Weak and electromagnetic forces are unified by electroweak interactions of
$SU\!(2)_L$ and $U\!(1)_Y$, whose
four bosons $W^1$, $W^2$, $W^3$, and $B$ respectively would be massless
if the symmetry were unbroken,
but it is `spontaneously' broken with the
Higgs mechanism~\cite{
higgs1964broken,
englert1964broken
}%
\footnote{%
Historical accuracy prefers `Brout–Englert–Higgs mechanism', but to
properly credit associated work, Peter~Higgs has reportedly suggested
``ABEGHHKtH mechanism''~\cite{close2011infinity}.%
}

\TODO{L and Y, charge Q = T3 - Y or whatever it is}

without the Higgs mechanism, which inserts four scalar fields and
`spontaneously breaks' the gauge symmetry with a potential that moves the
ground state of the model to a gauge-asymmetric configuration.
From this new ground state, linear combinations of the electroweak fields
acquire masses while ``eating up'' three of the four scalar fields to yield
the three weak bosons and the photon:
$(W^1, W^2)$ mix into $(W^+, W^-)$, and $(W^3, B)$ mix into $(Z, \gamma)$.
The fourth and final scalar field is the Higgs boson $h$~\cite{
rubakov1999classical,
cottingham2007greenwood
}.

@@

@@

\section{The \atlas\ Experiment}
\label{sec:context_atlas}
Towards explaining the main result of this thesis,
a search for effects from supersymmetric models
in data from the \atlas\ detector
on the Large Hadron Collider (LHC)~\cite{
atlas2022searches,
atlas2008experiment,
lhc2008machine
},
we first cover some relevant facts about the design and motivations
behind \atlas.

### 2022-09-27 main/chapter_searches.tex
some improvements to SM; some misc tweaks

Coin tosses, for example, can be imagined as random.
Earth's human population could toss a few billion coins to precisely
establish a probability of landing heads from the ratio of
$n$ (number of heads) to $N$ (number of tosses).
But suppose you are given a red envelope containing a coin;
which way up is the coin in the envelope?
You could guess or bet on the answer, and that decision might involve a tacitly
assumed probability.
A frequency, however, would only be permitted if the envelope had been packaged
at `random'.
Was the coin first tossed to randomize its face, or placed deliberately?
You do not know.
In the words of Jerzy~Neyman,
``I have put the word ``~random~'' in inverted commas because it is very
difficult to define what is meant by it in practice''~\cite{
Neyman1937Outline
}.

@@

\subsection{Fisher}
\label{sec:searches_fisher}
% likelihood
Fisher coined the term `likelihood'~\cite{
fisher1912fitting,
fisher1915frequency,
fisher1921probable,
fisher1922estimators
} in the usage that
we employ for the probability of observed data given varied
contexts,%
\footnote{%
Fisher used the term `likelihood' for the function exploring
contexts and `probability' for the distribution over
outcomes~\cite{fisher1925smrw}.
We choose the alternative word `prior' for his `probability' since the
likelihood is also a probability, just not a probability \emph{distribution}.
}
and deserves credit for popularizing its use in interpreting and communicating
data without the confounding influence of prior probabilities.
In his own words,%
\begin{quote}
\small
``%
We can know nothing of the probability of hypotheses or hypothetical
quantities.
On the other hand we may ascertain the likelihood of hypotheses
and hypothetical quantities by calculation from observations%
''~\cite{fisher1921probable},
\end{quote}
and furthermore we agree that in comparison to other prescriptions
(specifically confidence intervals, to be introduced in
Section~\ref{sec:searches_np}),
\begin{quote}
\small
``%
For all purposes, and more particularly for the \textit{communication} of the
relevant evidence supplied by a body of data, the values of the Mathematical
Likelihood are better fitted to analyse, summarize, and communicate
statistical evidence of types too weak to supply true probability statements%
''~\cite{fisher1956statistical}.
\end{quote}
For example, Figure~\ref{fig:searches_fisher_likelihood} displays a figure by
Fisher~\cite{fisher1956statistical} that illustrates how data can be usefully
reported by the likelihood's maximum and the regions in which it exceeds
certain ratios relative to that maximum.%
\footnote{%
We use such $L(\check{x})\pm\mathrm{ratio}$ visualizations for
$\pm1\sigma$ and $\pm2\sigma$ error bars on Poisson data in
Figure~7 of~\cite{lester2022hunting}.
When recommending error bar recipes, \atlas' statistics committee appears
to have not considered this option~\cite{cowan2011errorbars}.
% , although
% our statistics experts are surely familiar with Fisher's work.
}
That the likelihood contains all relevant information in the data is true by
definition in our formulation of Equation~\ref{eqn:searches_log_odds},
but non-Bayesian arguments find the same conclusion~\cite{
birnbaum1962foundations,
savage1962foundations
}.

>>>>

Coin tosses, for example, can be imagined as random.
Earth's human population could toss a few billion coins to precisely
establish a probability of landing heads from the ratio of
$n$ (number of heads) to $N$ (number of tosses).
But suppose you are given a red envelope containing a coin;
which way up is the coin in the envelope?
You could guess or bet on the answer, and that decision might involve a tacitly
assumed probability.
A frequency, however, would only be permitted if the envelope had been packaged
at `random'.
Was the coin first tossed to randomize its face, or placed deliberately?
You do not know.
In the words of Jerzy~Neyman,
``I have put the word ``\,random\,'' in inverted commas because it is very
difficult to define what is meant by it in practice''~\cite{
Neyman1937Outline
}.

@@

\subsection{Fisher}
\label{sec:searches_fisher}
% likelihood
Fisher coined the term `likelihood'~\cite{
fisher1912fitting,
fisher1915frequency,
fisher1921probable,
fisher1922estimators
} in the usage that
we employ for the probability of observed data given varied
contexts,%
\footnote{%
Fisher used the term `likelihood' for the function exploring
contexts and `probability' for the distribution over
outcomes~\cite{fisher1925smrw}.
We choose the alternative word `prior' for his `probability' since the
likelihood is also a probability, just not a probability \emph{distribution}.
}
and deserves credit for popularizing its use in interpreting and communicating
data without the confounding influence of prior probabilities.
In his own words,%
\begin{quote}
\small
``%
We can know nothing of the probability of hypotheses or hypothetical
quantities.
On the other hand we may ascertain the likelihood of hypotheses
and hypothetical quantities by calculation from observations%
''~\cite{fisher1921probable},
\end{quote}
and furthermore we agree that in comparison to other prescriptions
(specifically confidence intervals, to be introduced in
Section~\ref{sec:searches_np}),
\begin{quote}
\small
``%
For all purposes, and more particularly for the \textit{communication} of the
relevant evidence supplied by a body of data, the values of the Mathematical
Likelihood are better fitted to analyse, summarize, and communicate
statistical evidence of types too weak to supply true probability statements%
''~\cite{fisher1956statistical}.
\end{quote}
For example, Figure~\ref{fig:searches_fisher_likelihood} displays a figure by
Fisher~\cite{fisher1956statistical} that illustrates how data can be usefully
reported by the likelihood's maximum and the regions in which it exceeds
certain ratios relative to that maximum.%
\footnote{%
We use such $L(\check{x})\pm\mathrm{ratio}$ visualizations for
$\pm1\sigma$ and $\pm2\sigma$ error bars on Poisson data in
Figure~7 of~\cite{lester2022hunting}.
When recommending error bar recipes, \atlas's statistics committee appears
to have not considered this option~\cite{cowan2011errorbars}.
% , although
% our statistics experts are surely familiar with Fisher's work.
}
That the likelihood contains all relevant information in the data is true by
definition in our formulation of Equation~\ref{eqn:searches_log_odds},
but non-Bayesian arguments find the same conclusion~\cite{
birnbaum1962foundations,
savage1962foundations
}.

### 2022-09-27 main/draft.tex
some more text on SM, minor tweaks
some improvements to SM; some misc tweaks

% significance measures
ATLAS' recommended significance measure~\cite{atlas_significance} is
\begin{align}
\label{eqn:significance_atlas}
S_\mathrm{\atlas}(n; \mu, \sigma) =~&
\sqrt{2} \times
\mathrm{sign}(n - \mu) \times
\\[0.2em] \nonumber
&
\sqrt{
n\log\left(\frac{n(\mu + \sigma^2)}{\mu^2 + n\sigma^2}\right)
- \frac{\mu^2}{\sigma^2}\log\left(
1 + \frac{\sigma^2(n - \mu)}{\mu(\mu + \sigma^2)}
\right)
}
,
\end{align}
taken in a limit where $0\log(0x) = 0$ for the $n=0$ case.

@@

>>>>

% significance measures
ATLAS's recommended significance measure~\cite{atlas_significance} is
\begin{align}
\label{eqn:significance_atlas}
S_\mathrm{\atlas}(n; \mu, \sigma) =~&
\sqrt{2} \times
\mathrm{sign}(n - \mu) \times
\\[0.2em] \nonumber
&
\sqrt{
n\log\left(\frac{n(\mu + \sigma^2)}{\mu^2 + n\sigma^2}\right)
- \frac{\mu^2}{\sigma^2}\log\left(
1 + \frac{\sigma^2(n - \mu)}{\mu(\mu + \sigma^2)}
\right)
}
,
\end{align}
taken in a limit where $0\log(0x) = 0$ for the $n=0$ case.

@@

These are classified into the strong sector
% quarks
\begin{itemize}
\item up-type: up, charm, top ($u$, $c$, $t$), and
\item down-type: down, strange, bottom ($d$, $s$, $b$),
\end{itemize}
% leptons

% ex quotes

\begin{singlespacing}
\section{The Standard Model and beyond}
\label{sec:context_sm_and_beyond}
%
\begin{epigraphs}
\qitem{%
In fact, it is a defensible conjecture that no
theory can be complete and inviolable that emerges from finitely equipped minds
and survives finitely scoped experiment, which are the twin rickety foundations
for all theories.%
}%
{James~D.~Wells,
\textit{Discovery Beyond the Standard Model of~Elementary~Particle~Physics},
2019~\cite{wells2020discovery}}
\end{epigraphs}
\end{singlespacing}

\begin{singlespacing}
\section{The \atlas\ Experiment}
\label{sec:context_atlas}
%
\begin{epigraphs}
\qitem{%
if observing outer space gives us a view of the past, observing inner space
would surely give us a glimpse into the future - would be interesting if NASA
made a telescope for that%
}%
{Ken~M,
\textit{Comment: Yahoo! News},
2012~\cite{kenm2012inner}}
\end{epigraphs}
\end{singlespacing}

% \qitem{%
% One geometry can not be more true than another; it can only be more convenient.
% Geometry is not true, it is advantageous.}%
% {Robert~M.~Pirsig,
% \textit{Zen and the Art of Motorcycle Maintenance},
% 1974~\cite{pirsig1999zen}}
